---
title: "Troubleshoot_convergence"
output: html_document
date: "2025-02-28"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(tidyverse)
library(here)
library(MARSS)

load(here("shinyapp/data/sar_raw_data.rda"))
```

Are you are having trouble getting the model to converge, then start by
addressing the following:

```{r }
data_train<-sar_dart %>% 
  filter(between(year, 2002,2021))
```

1)  Are you trying to fit a bad model, e.g., fitting a non-stationary
    model to stationary data or fitting a model that specifies
    independence of errors or states to data that clearly violate that
    assumption or fitting a model that implies a particular stationary
    distribution to data that strongly violate that?
    -   Data do appear mostly stationary for DART

```{r}
ggplot(sar_raw_data %>% filter(index == "CUI") %>% drop_na(), aes(x = year, y = logit.s, color = sar.method)) +
  geom_line() +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "All available data")

#lm
ggplot(sar_raw_data %>% filter(index == "CUI", 
                               ifelse(sar.method == "DART", between(year, 2002,2021),TRUE)
                               ) %>% drop_na(), 
       aes(x = year, y = logit.s, color = sar.method, fill = sar.method))+
  geom_line() +
  geom_point()+
  geom_smooth(method = "lm", se = TRUE) +
    labs(title = "DART: 2002:2021")
#loess smoothing
ggplot(sar_raw_data %>% filter(index == "CUI", 
                               ifelse(sar.method == "DART", between(year, 2002,2021),TRUE)
                               ) %>% drop_na(), 
       aes(x = year, y = logit.s, color = sar.method, fill = sar.method))+
  geom_line() +
  geom_point()+
  geom_smooth(method = "loess", se = TRUE, span = .95) +
    labs(title = "DART: 2002:2021")

```



2)  Do you have confoundedparameters, e.g., two parameters that have the
    same effect (e.g., effectively two intercepts)?,
3)  Are you trying to fit a model to 1 data point somewhere, e.g., in a
    big multivariate dataset with lots of missing values?
    - no
4)  How many parameters are you trying to estimate per data point?
5)  Check your residuals (look at the QQplots in plot(fit)) for
    normality.
    
6)  Did you do any data transformations that would cause one of the
    variances to go to zero? Replacing 0s with a constant will do that.
    Try replacing them with NAs (missing). Do you have long strings of
    constant numbers in your data? Binned data often look like that, and
    that will drive Q to 0.

-------
```{r}
control_list<-list(maxit = 1000, allow.degen = FALSE)
```


# DART
- only data from 2002-2021
- does not include a covariate
- Next step, try pulling in SAR by 1) transport and 2) different reaches
```{r}
sar_dart<- sar_raw_data %>%
  filter(
    between(year,2002,2021),
    index == "CUI",
    sar.method == "DART"
  ) 


years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
m <- 1   
B <- diag(m)                       
U <- matrix(0, nrow = m, ncol = 1)  
Q <- "diagonal and unequal"         
A <- matrix(0)               
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, A = A, R = R)

fit1 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)

plot(fit1, silent = TRUE, plot.type = c("fitted.ytT", "xtT"))

```

# Scheuerell and Williams (2005)

## CUI
- data years: 1964 - 2005
- Q matrix diagonal and unequal
```{r }
sar_sw_cui<- sar_raw_data %>%
  filter(
    index == "CUI",
    sar.method == "Scheuerell and Williams (2005)"
  ) %>% 
  drop_na()


years <- sar_sw_cui$year
TT <- length(years)
dat <- matrix(sar_sw_cui$logit.s, nrow = 1)
index <- sar_sw_cui$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1)
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "diagonal and unequal"
A <- matrix(0)
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit2 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
plot(fit2, silent = TRUE) # plot.type = c("fitted.ytT", "xtT")
```

## CUTI
- data years: 1988 - 2005
- Q matrix unconstrained
```{r }
sar_sw_cuti<- sar_raw_data %>%
  filter(
    index == "CUTI",
    sar.method == "Scheuerell and Williams (2005)"
  ) %>% 
  drop_na()


years <- sar_sw_cuti$year
TT <- length(years)
dat <- matrix(sar_sw_cuti$logit.s, nrow = 1)
index <- sar_sw_cuti$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1) #changing to estimate "u" changes X2 to negative linear line
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z     
Q <- "unconstrained"
A <- matrix(0)
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit3 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list, control = control_list)
autoplot(fit3, silent = TRUE)
autoplot(fit3, silent = TRUE, plot.type = "fitted.ytt1")#one-step ahead

MARSSparamCIs(fit3)

#attempted to run without z-score for CUTI, did converge quicker (145 iterations) and some difference in R and Q output but model lower AIC and log-likelihood for z-scored covariate. 
```



# CJS
- next steps:
  1) rerun with updated CJS data (confirm or check with MC that this is all data not limited by length data)
  2) Include transport and/or temperature as covariate.
  3) rerun to calculate SAR for different reaches: BON-BOA, TWX-BOA, and LGR-LGA

## CUI
```{r}
sar_cjs<- sar_raw_data %>%
  filter(index == "CUI",sar.method == "CJS") %>%
  drop_na()

years <- sar_cjs$year
TT <- length(years)
dat <- matrix(sar_cjs$logit.s, nrow = 1)
index <- sar_cjs$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1) #changing to estimate "u" changes X2 to negative linear line
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "diagonal and unequal"#"unconstrained"
A <- matrix(0)
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit4 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
autoplot(fit4, silent = TRUE)
autoplot(fit4, silent = TRUE, plot.type = "fitted.ytt1")#one-step ahead

MARSSparamCIs(fit4)

```

## CUTI
- Q matrix unconstrained
```{r}
sar_cjs<- sar_raw_data %>%
  filter(index == "CUTI",sar.method == "CJS") %>%
  drop_na()

years <- sar_cjs$year
TT <- length(years)
dat <- matrix(sar_cjs$logit.s, nrow = 1)
index <- sar_cjs$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1) #changing to estimate "u" changes X2 to negative linear line
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z      
Q <- "unconstrained"
A <- matrix(0)
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit5 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
autoplot(fit5, silent = TRUE)
autoplot(fit5, silent = TRUE, plot.type = "fitted.ytt1")#one-step ahead

MARSSparamCIs(fit5)

```

## NCBI
- Q matrix unconstrained
- model process and observation error is explained a bit more with running ncbi as reported values without z-score
```{r}
sar_cjs<- sar_raw_data %>%
  filter(index == "NCBI",sar.method == "CJS") %>%
  drop_na()

years <- sar_cjs$year
TT <- length(years)
dat <- matrix(sar_cjs$logit.s, nrow = 1)
index <- sar_cjs$value
# index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1) #run as reported
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1) #changing to estimate "u" changes X2 to negative linear line
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index     
Q <- "unconstrained"
A <- matrix(0)
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit6 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
autoplot(fit6, silent = TRUE)
autoplot(fit6, silent = TRUE, plot.type = "fitted.ytt1")#one-step ahead

MARSSparamCIs(fit6)

```

## ICPB
- does not run with unconstrained Q
- does not converge if using z-scored value or unstandardized. Seems to be on a logit-scale
```{r}
sar_cjs<- sar_raw_data %>%
  filter(index == "ICPB",sar.method == "CJS") %>%
  drop_na()

years <- sar_cjs$year
TT <- length(years)
dat <- matrix(sar_cjs$logit.s, nrow = 1)
index <- sar_cjs$value
# index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1) #changing to estimate "u" changes X2 to negative linear line
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index    
Q <- "unconstrained"
A <- matrix(0)
R <- matrix("r")

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit7 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
autoplot(fit7, silent = TRUE)
autoplot(fit7, silent = TRUE, plot.type = "fitted.ytt1")#one-step ahead

MARSSparamCIs(fit6)

```

