---
title: "Troubleshoot_convergence"
output: html_document
date: "2025-02-28"
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
library(tidyverse)
library(here)
library(MARSS)

load(here("shinyapp/data/sar_raw_data.rda"))
```

Are you are having trouble getting the model to converge, then start by
addressing the following:

```{r }
data_train<-sar_dart %>% 
  filter(between(year, 2002,2021))
```

1)  Are you trying to fit a bad model, e.g., fitting a non-stationary
    model to stationary data or fitting a model that specifies
    independence of errors or states to data that clearly violate that
    assumption or fitting a model that implies a particular stationary
    distribution to data that strongly violate that?
    -   Data do appear mostly stationary for DART

```{r}
ggplot(sar_raw_data %>% filter(index == "CUI") %>% drop_na(), aes(x = year, y = logit.s, color = sar.method)) +
  geom_line() +
  geom_point()+
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "All available data")

#lm
ggplot(sar_raw_data %>% filter(index == "CUI", 
                               ifelse(sar.method == "DART", between(year, 2002,2021),TRUE)
                               ) %>% drop_na(), 
       aes(x = year, y = logit.s, color = sar.method, fill = sar.method))+
  geom_line() +
  geom_point()+
  geom_smooth(method = "lm", se = TRUE) +
    labs(title = "DART: 2002:2021")
#loess smoothing
ggplot(sar_raw_data %>% filter(index == "CUI", 
                               ifelse(sar.method == "DART", between(year, 2002,2021),TRUE)
                               ) %>% drop_na(), 
       aes(x = year, y = logit.s, color = sar.method, fill = sar.method))+
  geom_line() +
  geom_point()+
  geom_smooth(method = "loess", se = TRUE, span = .95) +
    labs(title = "DART: 2002:2021")

```



2)  Do you have confoundedparameters, e.g., two parameters that have the
    same effect (e.g., effectively two intercepts)?,
3)  Are you trying to fit a model to 1 data point somewhere, e.g., in a
    big multivariate dataset with lots of missing values?
    - no
4)  How many parameters are you trying to estimate per data point?
5)  Check your residuals (look at the QQplots in plot(fit)) for
    normality.
    
6)  Did you do any data transformations that would cause one of the
    variances to go to zero? Replacing 0s with a constant will do that.
    Try replacing them with NAs (missing). Do you have long strings of
    constant numbers in your data? Binned data often look like that, and
    that will drive Q to 0.

-------
```{r}
control_list<-list(maxit = 1000, allow.degen = FALSE, control = list(conv.test.slope.tol = 0.1)) 

#use to test convergence  at lower slope abstol
```

```{r CUI}
library(rerddap) 

### ERDAPP, NOAA upwelling indice CUI
# Set the URL of the ERDDAP server
url <- "https://upwell.pfeg.noaa.gov/erddap/"

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdUI45mo", url = url)
info # copy max range value from metadata

# Download the dataset
dat <- griddap(info,
               time = c("1964-01-01T00:00:00Z", "2025-01-15T00:00:00Z"), #currently pulling all data to date (adjust as needed)
               latitude = c(45,45)
)

# format time and lat/long
df_cui_raw<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day), 
         adjlong = longitude-360) #set to -180 to 180

## only include april CUI from ERDAPP data (for now)
df_cui<-df_cui_raw %>% 
  filter(month == 4) %>% 
  select(year, upwelling_index) %>% 
    rename("value" = upwelling_index)  %>%
  mutate(index = "CUI")
```


```{r CUTI}

# Get information about the erdCUTI dataset
info <- info(datasetid = "erdCUTImonthly", url = url)
info #copy max time range from metadata

# Download the dataset
dat <- griddap(info,
               time = c("1988-01-15T00:00:00Z", "2025-02-15T00:00:00Z"), #currently pulling all data to date
               latitude = c(45,45)
)


df_cuti<-dat$data %>% 
  mutate(time.day = lubridate::ymd_hms(time),
         year = year(time.day),
         month = month(time.day),
         day = day(time.day)) %>% 
  filter(month == 4) %>% 
  select(year, CUTI)  %>% 
  rename(value = "CUTI") %>% 
  mutate(index = "CUTI")
```
```{r NCBI}
library(readxl) #
#Northern copepod biomass data
#NCBI<-read_csv(here("data-raw", "northern_copepod_biomass_newportOR.csv")) 

url<-"https://www.fisheries.noaa.gov/s3/2023-10/TempOxyNoSo-10232023.xlsx"

# Download the file to a temporary location
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, temp_file, mode = "wb")

ncbi_raw<-read_excel(temp_file)


# Currently set to only pull April data but NOAA stoplight chart returns value for May - September. Future Research direction is attempting various month iterations.
df_ncbi<-ncbi_raw %>% 
  filter(Month == 4) %>% #select only april for now-- update with more research
  group_by(Year) %>% 
  summarise(value = mean(NORTHERN)) %>% 
  select("year" = Year, value) %>% 
  mutate(index = "NCBI")
```

```{r load_ICPB_data}
#Northern copepod biomass data (in-house)
# df_ICPB<-read_csv(here("data-raw", "IndexOfCoastalPreyBiomass.csv")) 
 
#to pull directly from stoplight table 
library(data.table)
year <- 2024 #update most recent year
url <- paste0("https://www.fisheries.noaa.gov/s3/2024-12/stoplight-raw-data-",year,".csv")
raw_data<-fread(url, header = TRUE, check.names = FALSE)

df_icpb<-raw_data %>% 
  mutate(across(-1, as.numeric)) %>%
  pivot_longer(-1, names_to = "year", values_to = "value") %>% 
  filter(`Ecosystem Indicators` == "Nearshore Ichthyoplankton\nLog(mg C 1,000 m-3; Jan-Mar)") %>% 
  select(year, value) %>% 
  mutate(index = "ICPB") %>% 
  mutate(year = as.integer(year))

range(df_icpb$value)
```

# DART 2000 -2022
### BON - BOA
```{r}
bon_dart<-read.csv(here("data/dart_BON_BOA_sar.csv")) %>% 
  drop_na(meanSAR) %>% 
  mutate(logit.s = qlogis(meanSAR/100)) %>% 
  select(year, logit.s) %>% 
  mutate(year = as.numeric(year))

bon_dart<-bon_dart %>% 
  left_join(df_cui, by = "year") %>% 
  mutate(sar.method = "DART")

sar_dart<- bon_dart %>% 
  left_join(df_cuti, by = "year")

```

#### no covariate
```{r}

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
m <- 1   
B <-  matrix("m", nrow = m, ncol = 1)                      
U <- matrix("u", nrow = m, ncol = 1)  
Q <- "equalvarcov" #doesn't really apply to 1 state variable         
A <- matrix(0)               
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}

autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
```
#### CUI
Attempt with set R to 0.05, Q to "unconstrained", U to be estimated and B estimated but equal
```{r}
{
years <- bon_dart$year
TT <- length(years)
dat <- matrix(bon_dart$logit.s, nrow = 1)
index <- bon_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #diagonal and unequal" #diag(m) #"identity" #"unconstrained"
U <-  matrix("u", nrow = m, ncol = 1) #"zero" #changing to  -1 to 1  works with varyin deg changes. higher(+ or -) values improves R.r and less in Q but estimate is worse.
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit6 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}
autoplot(fit6, silent = TRUE) # plot with all data (past and future observations) included in estimate
autoplot(fit6, silent = TRUE, plot.type = "fitted.ytt1")# plot with only past observations included in the estimate (one-step ahead)
# fitted.ytt1 is more realisitic of what real forecast would be-- but the extra tight CI in ytT likely means that the model is relying heavily on future observations and that the model  likely needs to be tweaked to better estimate B, U, Z, Q and R
MARSSparamCIs(fit6)
```

#### CUTI
```{r}
{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m) #"identity" #"unconstrained" "diagonal and unequal" #
U <-  matrix(0, nrow = m, ncol = 1) #"zero" 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit6)
```
#### NCBI
```{r}
sar_dart<-sar_raw_data %>% 
  filter(index == "NCBI", sar.method == "DART") %>% 
  select(year, value, index) %>% 
  right_join(select(bon_dart, year, logit.s), by = "year")

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m) #"identity" #"unconstrained" "diagonal and unequal" #
U <-  matrix(0, nrow = m, ncol = 1) #"zero" 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
#### ICPB
```{r}
sar_dart<-sar_raw_data %>% 
  filter(index == "ICPB", sar.method == "DART") %>% 
  select(year, value, index) %>% 
  right_join(select(bon_dart, year, logit.s), by = "year")

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-"diagonal and unequal" # diag(m) #"identity" #"unconstrained" 
U <-  matrix("u", nrow = m, ncol = 1) #"zero" 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```


### LGR - LGA

#### no covariate
```{r}
sar_dart<- sar_raw_data %>%
  filter(
    index == "CUI",
    sar.method == "DART"
  ) 

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
m <- 1   
B <-  matrix("m", nrow = m, ncol = 1)                      
U <- matrix("u", nrow = m, ncol = 1)  
Q <- "equalvarcov" #doesn't really apply to 1 state variable         
A <- matrix(0)               
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}

autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
```


#### CUI

```{r}
sar_dart<- sar_raw_data %>%
  filter(
    # between(year,2002,2021),
    index == "CUI",
    sar.method == "DART"
  ) %>% 
  drop_na

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  "diagonal and unequal" #diag(m) #"identity" #"unconstrained"
U <-  matrix(0, nrow = m, ncol = 1) #"zero" #changing to  -1 to 1  works with varyin deg changes. higher(+ or -) values improves R.r and less in Q but estimate is worse.
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit6 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}
DARTautoplot(fit6, silent = TRUE) # plot with all data (past and future observations) included in estimate
autoplot(fit6, silent = TRUE, plot.type = "fitted.ytt1")# plot with only past observations included in the estimate (one-step ahead)
# fitted.ytt1 is more realisitic of what real forecast would be-- but the extra tight CI in ytT likely means that the model is relying heavily on future observations and that the model  likely needs to be tweaked to better estimate B, U, Z, Q and R
MARSSparamCIs(fit6)
```

#### CUTI
```{r}
sar_dart<- sar_raw_data %>%
  filter(
    # between(year,2002,2021),
    index == "CUTI",
    sar.method == "DART"
  ) %>% 
  drop_na

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"identity" #"unconstrained"
U <-  matrix(0, nrow = m, ncol = 1) #"zero" 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}
autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
#### NCBI
```{r}
sar_dart<- sar_raw_data %>%
  filter(
    # between(year,2002,2021),
    index == "NCBI",
    sar.method == "DART"
  ) %>% 
  drop_na

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"identity" #"unconstrained"
U <-  matrix(0, nrow = m, ncol = 1) #"zero" 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <-  "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}

autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
#### ICPB
```{r}
sar_dart<- sar_raw_data %>%
  filter(
    # between(year,2002,2021),
    index == "ICPB",
    sar.method == "DART"
  ) %>% 
  drop_na

{
years <- sar_dart$year
TT <- length(years)
dat <- matrix(sar_dart$logit.s, nrow = 1)
index <- sar_dart$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- "diagonal and unequal" #diag(m) #"identity" #"unconstrained"
U <-  matrix("u", nrow = m, ncol = 1) #"zero" 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <-  "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list) #control = list(conv.test.slope.tol = 0.1) checked
}

autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```



# Scheuerell and Williams (2005)

## CUI
- data years: 1964 - 2005
- Q matrix diagonal and unequal
```{r }
sar_sw_cui<- sar_raw_data %>%
  filter(
    index == "CUI",
    sar.method == "Scheuerell and Williams (2005)"
  ) %>% 
  drop_na()

{
years <- sar_sw_cui$year
TT <- length(years)
dat <- matrix(sar_sw_cui$logit.s, nrow = 1)
index <- sar_sw_cui$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m)
U <- matrix(0, nrow = m, ncol = 1)
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "diagonal and unequal"
A <- matrix(0)
R <- matrix("r")
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit2 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
plot(fit2, silent = TRUE) # plot.type = c("fitted.ytT", "xtT")
```

## CUTI
- data years: 1988 - 2005
- Q matrix unconstrained
```{r }
sar_sw_cuti<- sar_raw_data %>%
  filter(
    index == "CUTI",
    sar.method == "Scheuerell and Williams (2005)"
  ) %>% 
  drop_na()

{
years <- sar_sw_cuti$year
TT <- length(years)
dat <- matrix(sar_sw_cuti$logit.s, nrow = 1)
index <- sar_sw_cuti$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- "diagonal and unequal"
U <- matrix("u", nrow = m, ncol = 1) #changing to estimate "u" changes X2 to negative linear line
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z     
Q <- "unconstrained"
A <- matrix(0)
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit3 <- MARSS::MARSS(dat, inits = inits_list, model = mod_list, control = control_list)
}
autoplot(fit3, silent = TRUE)
diagautoplot(fit3, silent = TRUE, plot.type = "fitted.ytt1")#one-step ahead

MARSSparamCIs(fit3)

#attempted to run without z-score for CUTI, did converge quicker (145 iterations) and some difference in R and Q output but model lower AIC and log-likelihood for z-scored covariate. 
```


# CJS
- next steps:
  1) rerun with updated CJS data (confirm or check with MC that this is all data not limited by length data)
  2) Include transport and/or temperature as covariate.
  3) rerun to calculate SAR for different reaches: BON-BOA, TWX-BOA, and LGR-LGA
  
```{r}
bon_sar_raw<-read.csv(here("data/bon.boa.surv.yr.1994.2021_250309.csv"))
bon_sar_raw<-bon_sar_raw[-1]
lgr_sar_raw<-read.csv(here("data/lgr.lga.surv.yr.1994.2021_250309.csv"))
lgr_sar_raw<-lgr_sar_raw[-1]


bon_sar<-bon_sar_raw %>% 
  left_join(df_cui, by = "year") %>% 
  mutate(sar.method = "CJS",
         logit.s = qlogis(BON_BOA_SAR)) %>% 
  select(-BON_BOA_SAR)


lgr_sar<-lgr_sar_raw %>% 
  left_join(df_cui, by = "year") %>% 
  mutate(sar.method = "CJS",
         logit.s = qlogis(LGR_LGA_SAR)) %>% 
  select(-LGR_LGA_SAR)


plotly::ggplotly(
  ggplot() +
geom_line(data = bon_sar, aes(x = year, y= plogis(logit.s)), color = "red") +
  geom_line(data = lgr_sar, aes(x = year, y= plogis(logit.s)), color = "green") +
geom_line(data = sar_raw_data %>% filter(index == "CUI", sar.method == "CJS"), aes(x = year, y = plogis(logit.s))) 
)

ggplot(bon_sar, aes(x = year, y = value))+
  geom_line()
```


```{r shifting mean}
library(ggplot2)
library(dplyr)

##CUI
# Assuming bon_sar has columns 'year' and 'value'
# Calculate the means for 5-year and 10-year intervals
bon_sar_avg <- bon_sar %>%
  mutate(interval_5yr = (year - min(year)) %/% 5,
         interval_10yr = (year - min(year)) %/% 10)

mean_5yr <- bon_sar_avg %>%
  group_by(interval_5yr) %>%
  summarize(mean_value_5yr = mean(value),
            start_year_5yr = min(year),
            end_year_5yr = max(year))

mean_10yr <- bon_sar_avg %>%
  group_by(interval_10yr) %>%
  summarize(mean_value_10yr = mean(value),
            start_year_10yr = min(year),
            end_year_10yr = max(year))

# Plot the data with horizontal line segments for each mean
ggplot(bon_sar_avg, aes(x = year, y = value)) +
  geom_line() +
  geom_segment(data = mean_5yr, aes(x = start_year_5yr, xend = end_year_5yr, y = mean_value_5yr, yend = mean_value_5yr), color = "blue", linetype = "dashed") +
  geom_segment(data = mean_10yr, aes(x = start_year_10yr, xend = end_year_10yr, y = mean_value_10yr, yend = mean_value_10yr), color = "red", linetype = "dotted") +
  labs(title = "Shifting Mean Across Years _CUI",
       subtitle = "Blue dashed lines: 5-year mean, Red dotted lines: 10-year mean",
       x = "Year",
       y = "Value")

## survival rate
mean_5yr <- bon_sar_avg %>%
  group_by(interval_5yr) %>%
  summarize(mean_value_5yr = mean(logit.s),
            start_year_5yr = min(year),
            end_year_5yr = max(year))

mean_10yr <- bon_sar_avg %>%
  group_by(interval_10yr) %>%
  summarize(mean_value_10yr = mean(logit.s),
            start_year_10yr = min(year),
            end_year_10yr = max(year))

# Plot the data with horizontal line segments for each mean
ggplot(bon_sar_avg, aes(x = year, y = logit.s)) +
  geom_line() +
  geom_segment(data = mean_5yr, aes(x = start_year_5yr, xend = end_year_5yr, y = mean_value_5yr, yend = mean_value_5yr), color = "blue", linetype = "dashed") +
  geom_segment(data = mean_10yr, aes(x = start_year_10yr, xend = end_year_10yr, y = mean_value_10yr, yend = mean_value_10yr), color = "red", linetype = "dotted") +
  labs(title = "Shifting Mean Across Years _ SAR",
       subtitle = "Blue dashed lines: 5-year mean, Red dotted lines: 10-year mean",
       x = "Year",
       y = "SAR (logit)")



```


## BON - BOA
### no covariate
```{r}
sar<-bon_sar
{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
m <- 1   
B <-  matrix("m", nrow = m, ncol = 1)                      
U <- matrix("u", nrow = m, ncol = 1)  
Q <- "equalvarcov" #doesn't really apply to 1 state variable         
A <- matrix(0)               
R <- matrix("r")
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}

autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
```
### CUI
```{r}
{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  "diagonal and unequal" #diag(m) #"identity" #"unconstrained"
U <-  matrix(0, nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```

### CUTI

```{r}
sar<-sar %>% 
  select(year, logit.s) %>% 
  left_join(df_cuti, by = "year") 
{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  "diagonal and equal" #diag(m) 
U <-  matrix(0, nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```

### NCBI

```{r}
#started in 1997
sar<-sar %>% 
  select(year, logit.s) %>% 
  left_join(df_ncbi, by = "year") %>% 
  drop_na()

{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"diagonal and equal" #
U <-  matrix(0, nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
### ICPB

```{r}
#started in 1997
sar<-sar %>% 
  select(year, logit.s) %>% 
  left_join(df_icpb, by = "year") %>% 
  drop_na()

{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"diagonal and unequal" #
U <-  matrix(0, nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
## LGR - LGA
### no covariate
```{r}
sar<-lgr_sar
{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
m <- 1   
B <-  matrix("m", nrow = m, ncol = 1)                      
U <- matrix(0, nrow = m, ncol = 1)  
Q <- "equalvarcov" #doesn't really apply to 1 state variable         
A <- matrix(0)               
R <- matrix("r")
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}

autoplot(fit, silent = TRUE)
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
```
### CUI
```{r}
# model converges with CUI (+ adjustments)
# Estimation converged in 159 iterations. 
# Log-likelihood: -32.95369 
# AIC: 79.90738   AICc: 85.5073
{
years <- lgr_sar$year
TT <- length(years)
dat <- matrix(lgr_sar$logit.s, nrow = 1)
index <- lgr_sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <- diag(m) #"identity" #"unconstrained" #"diagonal and equal" #
U <-  matrix(0, nrow = m, ncol = 1) #"zero" #changing to  -1 to 1  works with varyin deg changes. higher(+ or -) values improves R.r and less in Q but estimate is worse.
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained"
A <- matrix(0)
R <-matrix(0.05) 

inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit<- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) # plot with all data (past and future observations) included in estimate
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")# plot with only past observations included in the estimate (one-step ahead)
# fitted.ytt1 is more realisitic of what real forecast would be-- but the extra tight CI in ytT likely means that the model is relying heavily on future observations and that the model  likely needs to be tweaked to better estimate B, U, Z, Q and R
MARSSparamCIs(fit)
```

### CUTI
```{r}
sar<-lgr_sar %>% 
  select(year, logit.s) %>% 
  left_join(df_cuti, by = "year") 


{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"diagonal and equal" #
U <-  matrix(0, nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.05)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
### NCBI

```{r}
#started in 1997
sar<-sar %>% 
  select(year, logit.s) %>% 
  left_join(df_ncbi, by = "year") %>% 
  drop_na()

{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"diagonal and equal" #
U <-  matrix("u", nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
### ICPB


```{r}
#started in 1997
sar<-sar %>% 
  select(year, logit.s) %>% 
  left_join(df_icpb, by = "year") %>% 
  drop_na()

{
years <- sar$year
TT <- length(years)
dat <- matrix(sar$logit.s, nrow = 1)
index <- sar$value
index_z <- matrix((index - mean(index)) / sqrt(var(index)), nrow = 1)
m<-2  
B <-  diag(m) #"diagonal and equal" #
U <-  matrix("u", nrow = m, ncol = 1) 
Z <- array(NA, c(1, m, TT))  
Z[1,1,] <- rep(1, TT)       
Z[1,2,] <- index_z       
Q <- "unconstrained" #"diagonal and unequal"
A <- matrix(0)
R <- matrix(.1)
inits_list <- list(x0 = matrix(c(0), nrow = m))
mod_list <- list(B = B, U = U, Q = Q, Z = Z, A = A, R = R)
fit <- MARSS::MARSS(dat, inits = inits_list, model = mod_list)
}
autoplot(fit, silent = TRUE) 
autoplot(fit, silent = TRUE, plot.type = "fitted.ytt1")
MARSSparamCIs(fit)
```
